Inferring the existence and direction of causal associations in the face of measurement error

### 2017-02-01 docs/manuscript.tex
updated

Inference of the causal relationship between two traits can be achieved
by combining genetic associations with a mediation-based approach, as is
done in the causal inference test (CIT) and others. We show here,
however, that measurement error in the phenotypes can lead to
mediation-based approaches inferring the wrong causal direction, and
that increasing sample sizes has the adverse effect of increasing
confidence in the wrong answer. Here we develop an extension to
Mendelian randomisation, a method that uses genetic associations in an
instrumentation framework, that enables inference of the causal
direction with some advantages. First, it is less susceptible to bias in
the presence of measurement error; second, it is more statistically
efficient; and third, it can be performed using only summary level data
from genome-wide association studies. We apply the method to infer the
causal direction between DNA methylation and gene expression levels. Our
results demonstrate that, in general, DNA methylation is more likely to
be the causal factor, but that such high level conclusions are
susceptible to differential measurement error between the platforms. We
emphasise that the use of several different methods is important to
triangulate reliable conclusions.

@@

Measurement error (or observational error, henceforth referred to simply
as measurement error) is the difference between the measured value of a
quantity and its true value. This study focuses specifically on
non-differential measurement error where all strata of a measured
variable have the same error rate, which can manifest as changes in
scale or levels of imprecision (noise) of the measurement. Such
variability can arise through a whole plethora of mechanisms, which are
often unique to the study design and difficult to
avoid\textsuperscript{7,8}. Array technology is now commonly used to
obtain high throughput phenotyping at low cost, but comes with the
problem of having imperfect resolution, for instance methylation levels
as measured by the Illumina450k chip are prone to have some amount of
noise around the true value due to imperfect
sensitivity\textsuperscript{9,10}. Relatedly, if the measurement of
biological interest is the methylation level in a T cell, then
measurement error of this value can be introduced by using methylation
levels from whole blood samples because the measured value will be an
assay of many cell types\textsuperscript{11}.

Measurement error will of course arise in more basic data too, for
example when measuring BMI one is typically interested in using this as
a proxy for adiposity, but it is clear that the correlation between BMI
and underlying adiposity is not perfect\textsuperscript{12}. A similar
problem of biological misspecification is unavoidable in disease
diagnosis, and measuring behaviour such as smoking or diet is
notoriously difficult to do accurately. Measurement error can also be
introduced after the data has been collected, for example the
transformation of non-normal data for the purpose of statistical
analysis will lead to a new variable that will typically incur both
changes in scale and imprecision (noise) compared to the original
variable. The sources of measurement error are not limited to this
list\textsuperscript{8}, and its impact has been explored in the
epidemiological literature extensively\textsuperscript{13,14}. Given the
near-ubiquitous presence of measurement error in phenomic data it is
vital to understand its impact on the tools we use for causal inference.

@@

By utilising genetic instruments in different ways, mediation-based
analysis and MR models have properties that confer some advantages and
some disadvantages for reliable causal inference. An important advantage
of mediation-based analysis is that the true underlying biology of the
genetic instrument doesn't necessarily need to be known as long as it
associates with the putative exposure. In the CIT framework (described
fully in the Methods) for example, the test statistic is different if
you test for the exposure causing the outcome or the outcome causing the
exposure, allowing the researcher to infer the direction of causality
between two variables by performing the test in both directions and
choosing the model with the strongest evidence. The CIT can also
distinguish between several putative models of association between the
traits and the SNP\textsuperscript{4,18}. Such is not the case for MR,
where in order to infer the direction of causality between two traits
the instrument must be known to influence the exposure primarily,
associating with the outcome only through the exposure.

Assuming biological knowledge of genetic associations can be problematic
because if there exists a putative association between two variables,
with the instrumental SNP being robustly associated with each, it can be
difficult to determine which of the two variables is subject to the
primary effect of the SNP (i.e.~for which of the two variables is the
SNP a valid instrument? Figure 1). By definition, we expect that if the
association is causal then a SNP for the exposure will be associated
with the outcome, such that if the researcher erroneously uses the SNP
as an instrument for the outcome then they are likely to see an
apparently robust causal association of outcome on exposure. Such a
situation can arise in many scenarios. Genome-wide association studies
(GWASs) that identify genetic associations for complex traits are, by
design, hypothesis free and agnostic of genomic function, and it often
takes years of follow up studies to understand the biological nature of
a putative GWAS hit\textsuperscript{27}. If multiple instruments are
available for an hypothesised exposure, which is increasingly typical
for complex traits that are analysed in large GWAS consortia, then
techniques can be applied to mitigate these issues\textsuperscript{16}.
But these techniques cannot always be applied in the case of determining
causal directions between 'omic measures where typically only one
cis-acting SNP is known. For example if a DNA methylation probe is
associated with expression of an adjacent gene, then is a cis-acting SNP
an instrument for the DNA methylation level, or the gene expression
level\textsuperscript{28} (Figure 1)?

MR does however have some important advantages over the mediation-based
approaches. The mediation-based approaches require that the exposure,
outcome and instrumental variables are all measured in the same data,
whereas recent extensions to MR circumvent this requirement, allowing
causal inference to be drawn when exposure variables and outcome
variables are measured in different samples\textsuperscript{29}. This
has the crucial advantage of improving statistical power by allowing
analysis in much larger sample sizes, and dramatically expands the
breadth of possible phenotypic relationships that can be
evaluated\textsuperscript{26}. Additionally, when MR assumptions are
satisfied the method is substantially more robust to there being
measurement error in the exposure variable. Indeed instrumental variable
(IV) analysis was in part initially introduced as a correction for
measurement error in the exposure\textsuperscript{30}, whereas it has
been noted that both classic mediation-based
analyses\textsuperscript{13,14,31,32} and mediation-based methods that
use instrumental variables\textsuperscript{33,34} are prone to be
unreliable in its presence.

Using theory and simulations we show how non-differential measurement
error in phenotypes can lead to unreliable causal inference in the
mediation-based CIT method. We then present an extension to MR that
allows researchers to ascertain the causal direction of an association
even when the biology of the instruments are not fully understood, and
also a method to evaluate the sensitivity of the result of this
extension to measurement error. This improves the utility of MR in cases
where mediation based methods might have otherwise been used
preferentially. Finally, we apply this new method to infer the direction
of causation between DNA methylation levels and gene expression levels.

@@

Note that the same correlation test approach can be applied to a
two-sample MR\textsuperscript{29} setting. Two-sample MR refers to the
case where the SNP-exposure association and SNP-outcome association are
calculated in different samples (e.g.~from publicly available summary
statistics). Here the Steiger test of two independent correlations can
be applied where.

@@

The Steiger test for inferring if \(x \rightarrow y\) is based on
evaluating \(\rho_{gx} > \rho_{gy}\). However, \(\rho_{gx}\) and
\(\rho_{gy}\) are underestimated if \(x\) and \(y\) are measured
imprecisely. If, for example, \(x\) has lower measurement precision than
\(y\) then we might empirically obtain \(\rho_{g,x_o} > \rho_{g,y_o}\).
As we show in the appendix, the maximum measurement error of \(x_o\) is
\(\rho_{g,x_o}\) and the minimum is 0 (same applies to \(y_o\)). It is
possible to simulate what the inferred causal direction would be for all
values within these bounds. To evaluate how reliable, \(R\), the Steiger
test is to potential measurement imprecision in \(x\) and \(y\) we
integrate over the simulated \(R\) values for the possible measurement
error values which are compatible with the direction of causality
obtained empirically from the Steiger test. We also do the same for the
measurement error values which are not compatible. The ratio of these
quantities is an indication of how much of the measurement error space
supports the inferred causal direction. A ratio \(R=1\) indicates that
the Steiger test is highly sensitive to measurement error, with equal
weight of the measurement error parameter space supporting both
directions of causality. Full details are provided in Appendix 2.

@@

Two sample MR\textsuperscript{29} was performed using the summary
statistics presented in\textsuperscript{28} for each of 4122
associations between DNA methylation and gene expression levels. To do
this we obtained a list of 458 gene expression - DNA methylation
associations as reported in Shakhbazov et al\textsuperscript{28}. These
were filtered to be located on the same chromosome, have robost
correlations after correcting for multiple testing, and to share a SNP
that had a robust cis-acting effect on both the DNA methylation probe
and the gene expression probe. Because only summary statistics were
available (effect, standard error, effect allele, sample size, p-values)
for the instrumental SNP on the methylation and gene expression levels,
the Steiger test of two independent correlations was used to infer the
direction of causality for each of the associations. The Wald ratio test
was then used to estimate the causal effect size for the estimated
direction for each association.

@@

\subsubsection{MR analysis of gene expression and DNA methylation
levels}\label{mr-analysis-of-gene-expression-and-dna-methylation-levels}

We used the Steiger test to infer the direction of causality between DNA
methylation levels and gene expression levels and found that the causal
direction commonly goes in both directions (Figure 6a). We then went on
to predict the causal directions of the associations for varying levels
of possible measurement error. Figure 6a shows that the qualitative
result of which of DNA methylation or gene expression is more commonly
the causal factor is very strongly determined by measurement error.

We performed two sample MR\textsuperscript{29} for each association in
the direction of causality that we estimated to be the correct direction
from the Steiger test. We observed that the sign of the MR estimate was
generally in the same direction as the Pearson correlation coefficient
reported by Shakhbazov et al\textsuperscript{28}, but with a few
exceptions (Figure 6b). There was a moderate correlation between the
absolute magnitudes of the causal correlation and the observational
Pearson correlation (r = 0.4538942). Together these inferences suggest
that even for relatively low level associations it is important to use
causal inference methods over observational associations to infer causal
effect sizes.

We also observed that for associations where methylation caused gene
expression the causal effect was more likely to be negative than for the
associations where gene expression caused methylation (OR = 0.6103961
(95\% CI 0.3622428 - 1.029752), Figure 6c), suggesting that reducing
gene expression levels at a controlling CpG typically leads to increased
gene expression levels, consistent with expectation\textsuperscript{40}.

@@

Researchers are often confronted with the problem of making causal
inferences using a statistical framework on observational data. In the
epidemiological literature issues of measurement error in mediation
analysis is relatively well explored\textsuperscript{41}. Our analysis
extends this to related methods such as CIT that are widely used in
predominantly 'omic data. These methods are indeed liable to the same
issues as standard mediation based analysis, and specifically we show
that as measurement error in the exposure variable increases, CIT is
likely to have reduced statistical power, and liable to infer the wrong
direction of causality. We also demonstrate that, though unintuitive,
increasing sample size does not resolve the issue, rather it leads to
more extreme p-values for the model that predicts the wrong direction of
causality.

Under many circumstances a practical solution to this problem is to use
Mendelian randomisation instead of methods such as the CIT or similar
that are based on mediation. Inferring the existence of causality using
Mendelian randomisation is robust in the face of measurement error and,
if the researcher has knowledge about the biology of the instrument
being used in the analysis, can offer a direct solution to the issues
that the CIT faces. This assumption is often reasonable, for example
SNPs are commonly used as instruments when they are found in genes with
known biological relevance for the trait of interest. But on many
occasions this is not the case, and it may be tempting to resort to
using methods based on mediation in order to be able to both ascertain
if there is a causal association and to infer the direction of
causality. Here we have described a simple extension to MR which can be
used as an alternative to mediation based methods. We show that this
method is still liable to measurement error, but because it has
different properties to the CIT it offers two main advantages. First, it
uses a formal statistical framework to test for the robostness of the
assumed direction of causality. Second, after testing in a comprehensive
range of scenarios the MR based approach is less likely to infer the
wrong direction of causality compared to CIT, while substantially
improving power over CIT in the cases where \(d > 0\).

We demonstrate this new method by evaluating the causal relationships of
458 known associations between DNA methylation and gene expression
levels using summary level data. The causal direction is heavily
influenced by how much measurement error is present in the different
assaying platforms. For example, if DNA methylation measures typically
have higher measuremet error than gene expression measures then our
analysis suggests that DNA methylation levels would be more often the
causal factor in the association. Indeed, previous studies which have
evaluated measurement error in these platforms do support this
position\textsuperscript{42,43}, though making strong conclusions for
this analysis is difficult because measurement error is likely to be
study specific.

In our simulations we focused on the simple case of a single instrument
in a single sample setting with a view to making a fair comparison
between MR and the various mediation-based methods available. However,
MR strategies exist for other scenarios, for example if there are
multiple instruments for the hypothesised exposure then the problems
described here may be circumvented\textsuperscript{16}. Likewise, if
there is at least one instrument for each trait then bi-directional MR
can offer solutions to inferring the causal
direction\textsuperscript{44}. In this work we assumed that pleiotropy
(the influence of the instrument on the outcome through a mechanism
other than the exposure) was not present. Recent method developments in
MR\textsuperscript{24,25} have focused on accounting for the issues that
horizontal pleiotropy can introduce, but how they perform in the
presence of measurement error remains to be explored. An important
advantage that MR confers over most mediation based analysis is that it
can be performed in two samples, which can considerably improve power
and expand the scope of analysis. However, whether there is a
substantive difference in two sample MR versus one sample MR in how
measurement error has an effect is not yet fully understood. We have
also assumed no measurement error in the genetic instrument, which is
not unreasonable given the strict QC protocols that ensure high quality
genotype data is available to most studies. We have restricted the scope
to only exploring non-differential measurement error and avoided the
complications incurred if measurement error in the exposure and outcome
is correlated, however our analysis goes beyond many previous
explorations of measurement error by assessing the impacts of both
imprecision (noise) and linear transformations of the true variable on
causal inference.

Mediation based network approaches, that go beyond analyses of two
variables, are very well established\textsuperscript{34} and have a
number of extensions that make them valuable tools, including for
example network construction. But because they are predicated on the
basic underlying principles of mediation they are liable to suffer from
the same issues of measurement error. Recent advances in MR methodology,
for example applying MR to genetical genomics\textsuperscript{45},
multivariate MR\textsuperscript{46} and mediation through
MR\textsuperscript{47,48} may offer more robust alternatives for these
more complicated problems.

@@

Figure 1: Gene expression levels (blue blocks) and DNA methylation
levels (green triangles) may be correlated but the causal direction is
unknown. If a SNP (yellow circle) is associated with both DNA
methylation and gene expression levels then it can be used as an
instrument, but there are three competing models for these variables.
The causal inference test (CIT) attempts to delineate between them. a)
Methylation causes gene expression. The left figure shows that the SNP
influences methylation levels that in turn influence gene expression
levels. The right figure shows the directed acyclic graph that
represents this model. Faded symbols represent the measured values
whereas solid symbols represent the true values. b) The same as in A,
except the causal direction is from gene expression to DNA methylation.
c) A model of confounding, where gene expression and DNA methylation are
not causally related, but the SNP influences them each through separate
pathways or a confounder.

@@

Figure 2: The CIT was performed on simulated variables where the
exposure influenced the outcome and the exposure was instrumented by a
SNP. The test statistic from CIT when testing if the exposure caused the
outcome (the true model) is in red, and the test for the outcome causing
the exposure (false model) is in green. Rows of plots represent the
sample sizes used for the simulations. As measurement error increases
(decreasing values on x-axis) the test statistic for the incorrect model
gets stronger and the test statistic for the correct model gets weaker.

@@

Figure 6: Using 458 putative associations between DNA methylation and
gene expression we used the Steiger test to infer the direction of
causality between them. a) The rightmost bar shows the proportion of
associations for each of the two possible causal directions (colour key)
assuming no measurement error in either gene expression or DNA
methylation levels. The proportions change when we assume different
levels of measurement error in gene expression levels (x-axis) or DNA
methylation levels (columns of boxes). If there is systematically higher
measurement error in one platform it will appear to be less likely to be
the causal factor. b) The relationship between the Pearson correlation
between DNA methylation and gene expression levels (x-axis) and the
causal estimate (scaled to be in standard deviation units, y-axis). c)
Distribution of estimated causal effect sizes, stratified into
associations inferred to be due to DNA methylation causing expression
(blue) and expression causing DNA methylation (red).

@@

\hypertarget{ref-Shakhbazov2016}{}
28. Shakhbazov, K. \emph{et al.} Shared genetic control of expression
and methylation in peripheral blood. \emph{BMC genomics} \textbf{17,}
278 (2016).

\hypertarget{ref-Pierce2013}{}
29. Pierce, B. L. \& Burgess, S. Efficient design for Mendelian
randomization studies: subsample and 2-sample instrumental variable
estimators. \emph{American journal of epidemiology} \textbf{178,}
1177--84 (2013).

\hypertarget{ref-Ashenfelter1994}{}
30. Ashenfelter, O. \& Krueger, A. B. Estimates of the Economic Return
to Schooling from a New Sample of Twins. \emph{The American Economic
Review} \textbf{84,} 1157--1173 (1994).

\hypertarget{ref-Nagarajan2013}{}
31. Nagarajan, R. \& Scutari, M. Impact of noise on molecular network
inference. \emph{PloS one} \textbf{8,} e80735 (2013).

\hypertarget{ref-Shpitser2010}{}
32. Shpitser, I., VanderWeele, T. \& Robins, J. On the validity of
covariate adjustment for estimating causal effects. \emph{Proceedings of
the Twenty Sixth Conference on Uncertainty in Artificial Intelligence
(UAI-10)} 527--536 (2010).

\hypertarget{ref-Wang2015}{}
33. Wang, L. \& Michoel, T. Detection of regulator genes and eQTLs in
gene networks. \emph{arXiv} \textbf{arXiv:1512,} (2015).

\hypertarget{ref-Lagani2015}{}
34. Lagani, V., Triantafillou, S., Ball, G., Tegner, J. \& Tsamardinos,
I. in \emph{Uncertainty in biology: A computational modeling approach}
47 (Springer, 2015). at
\textless{}\href{https://books.google.com/books?id=8SLUCgAAQBAJ\%7B/\&\%7Dpgis=1}{https://books.google.com/books?id=8SLUCgAAQBAJ\{\textbackslash{}\&\}pgis=1}\textgreater{}

@@

\hypertarget{ref-Bird2002}{}
40. Bird, A. DNA methylation patterns and epigenetic memory. \emph{Genes
\& development} \textbf{16,} 6--21 (2002).

>>>>

Inference of the causal structure that induces correlations between two
traits can be achieved by combining genetic associations with a
mediation-based approach, as is done in the causal inference test (CIT)
and others. However, we show that measurement error in the phenotypes
can lead to mediation-based approaches inferring the wrong causal
direction, and that increasing sample sizes has the adverse effect of
increasing confidence in the wrong answer. Here we introduce an
extension to Mendelian randomisation, a method that uses genetic
associations in an instrumentation framework, that enables inference of
the causal direction between traits, with some advantages. First, it is
less susceptible to bias in the presence of measurement error; second,
it is more statistically efficient; third, it can be performed using
only summary level data from genome-wide association studies; and
fourth, its sensitivity to measurement error can be evaluated. We apply
the method to infer the causal direction between DNA methylation and
gene expression levels. Our results demonstrate that, in general, DNA
methylation is more likely to be the causal factor, but this result is
highly susceptible to bias induced by systematic differences in
measurement error between the platforms. We emphasise that, where
possible, implementing MR alongside other approaches such as CIT is
important to triangulate reliable conclusions about causality.

@@

Measurement error is the difference between the measured value of a
quantity and its true value. This study focuses specifically on
non-differential measurement error where all strata of a measured
variable have the same error rate, which can manifest as changes in
scale or levels of imprecision (noise) of the measurement. Such
variability can arise through a whole plethora of mechanisms, which are
often unique to the study design and difficult to
avoid\textsuperscript{7,8}. Array technology is now commonly used to
obtain high throughput phenotyping at low cost, but comes with the
problem of having imperfect resolution, for instance methylation levels
as measured by the Illumina450k chip are prone to have some amount of
noise around the true value due to imperfect
sensitivity\textsuperscript{9,10}. Relatedly, if the measurement of
biological interest is the methylation level in a T cell, then
measurement error of this value can be introduced by using methylation
levels from whole blood samples because the measured value will be an
assay of many cell types\textsuperscript{11}.

Measurement error will of course arise in other types of data too, for
example when measuring BMI one is typically interested in using this as
a proxy for adiposity, but it is clear that the correlation between BMI
and underlying adiposity is not perfect\textsuperscript{12}. A similar
problem of biological misspecification is unavoidable in disease
diagnosis, and measuring behaviour such as smoking or diet is
notoriously difficult to do accurately. Measurement error can also be
introduced after the data has been collected, for example the
transformation of non-normal data for the purpose of statistical
analysis will lead to a new variable that will typically incur both
changes in scale and imprecision (noise) compared to the original
variable. The sources of measurement error are not limited to this
list\textsuperscript{8}, and its impact has been explored in the
epidemiological literature extensively\textsuperscript{13,14}. Given the
near-ubiquitous presence of measurement error in phenomic data it is
vital to understand its impact on the tools we use for causal inference.

@@

By utilising genetic instruments in different ways, mediation-based
analysis and MR models have properties that confer some advantages and
some disadvantages for reliable causal inference. An important advantage
of mediation-based analysis is that in order to infer the causal
direction the true underlying biology of the genetic instrument doesn't
necessarily need to be known as long as it associates with the putative
exposure. In the CIT framework (described fully in the Methods) for
example, the test statistic is different if you test for the exposure
causing the outcome or the outcome causing the exposure, allowing the
researcher to infer the direction of causality between two variables by
performing the test in both directions and choosing the model with the
strongest evidence. The CIT also has the valuable property of being able
to distinguish between several putative causal graphs that link the
traits with the SNP (Figure 1). Such is not the case for MR, where in
order to infer the direction of causality between two traits the
instrument must be known to influence the exposure primarily,
associating with the outcome only through the exposure.

Assuming biological knowledge of genetic associations can be problematic
because if there exists a putative association between two variables,
with the SNP being robustly associated with each, it can be difficult to
determine which of the two variables is subject to the primary effect of
the SNP (i.e.~for which of the two variables is the SNP a valid
instrument? Figure 1). By definition, we expect that if the association
is causal then a SNP for the exposure will be associated with the
outcome, such that if the researcher erroneously uses the SNP as an
instrument for the outcome then they are likely to see an apparently
robust causal association of outcome on exposure. Such a situation can
arise in many scenarios. Genome-wide association studies (GWASs) that
identify genetic associations for complex traits are, by design,
hypothesis free and agnostic of genomic function, and it often takes
years of follow up studies to understand the biological nature of a
putative GWAS hit\textsuperscript{27}. If multiple instruments are
available for an hypothesised exposure, which is increasingly typical
for complex traits that are analysed in large GWAS consortia, then
techniques can be applied to mitigate these issues\textsuperscript{16}.
But these techniques cannot always be applied in the case of determining
causal directions between 'omic measures where typically only one
cis-acting SNP is known. For example if a DNA methylation probe is
associated with expression of an adjacent gene, then is a cis-acting SNP
an instrument for the DNA methylation level, or the gene expression
level (Figure 1)?

MR does however have some important advantages over the mediation-based
approaches. The mediation-based approaches require that the exposure,
outcome and instrumental variables are all measured in the same data,
whereas recent extensions to MR circumvent this requirement, allowing
causal inference to be drawn when exposure variables and outcome
variables are measured in different samples\textsuperscript{28}. This
has the crucial advantage of improving statistical power by allowing
analysis in much larger sample sizes, and dramatically expands the
breadth of possible phenotypic relationships that can be
evaluated\textsuperscript{26}. Additionally, when MR assumptions are
satisfied the method is substantially more robust to there being
measurement error in the exposure variable. Indeed instrumental variable
(IV) analysis was in part initially introduced as a correction for
measurement error in the exposure\textsuperscript{29}, whereas it has
been noted that both classic mediation-based
analyses\textsuperscript{13,14,30,31} and mediation-based methods that
use instrumental variables\textsuperscript{32,33} are prone to be
unreliable in its presence.

Using theory and simulations we show how non-differential measurement
error in phenotypes can lead to unreliable causal inference in the
mediation-based CIT method. We then present an extension to MR that
allows researchers to ascertain the causal direction of an association
even when the biology of the instruments are not fully understood, and
also a method to evaluate the sensitivity of the result of this
extension to measurement error. This improves the utility of MR in cases
where mediation based methods might have otherwise been used
preferentially. Finally, we apply this new method to infer the direction
of causation between DNA methylation levels and gene expression levels.
Our analyses highlight that because these different causal inference
techniques have varying strengths and weaknesses, triangulation of
evidence from as many sources as possible should be practiced in causal
inference\textsuperscript{34}.

@@

Note that the same correlation test approach can be applied to a
two-sample MR\textsuperscript{28} setting. Two-sample MR refers to the
case where the SNP-exposure association and SNP-outcome association are
calculated in different samples (e.g.~from publicly available summary
statistics). Here the Steiger test of two independent correlations can
be applied where.

@@

The Steiger test for inferring if \(x \rightarrow y\) is based on
evaluating \(\rho_{gx} > \rho_{gy}\). However, \(\rho_{gx}\) and
\(\rho_{gy}\) are underestimated if \(x\) and \(y\) are measured
imprecisely. If, for example, \(x\) has lower measurement precision than
\(y\) then we might empirically obtain \(\rho_{g,x_o} > \rho_{g,y_o}\).
As we show in the appendix, the maximum measurement error of \(x_o\) is
\(\rho_{g,x_o}\) and the minimum is 0 (same applies to \(y_o\)). It is
possible to simulate what the inferred causal direction would be for all
values within these bounds. To evaluate how reliable, \(R\), the Steiger
test is to potential measurement imprecision in \(x\) and \(y\) we
integrate over the simulated \(R\) values for the possible measurement
error values which are compatible with the direction of causality
obtained empirically from the Steiger test. We also do the same for the
measurement error values which are not compatible. The ratio of these
quantities is an indication of how much of the measurement error space
supports the inferred causal direction. A ratio \(R=1\) indicates that
the Steiger test is highly sensitive to measurement error, with equal
weight of the measurement error parameter space supporting both
directions of causality. Full details are provided in Appendix 2 and
Figure 3a.

@@

Two sample MR\textsuperscript{28} was performed using the summary
statistics presented in\textsuperscript{40} for each of 4122
associations between DNA methylation and gene expression levels. To do
this we obtained a list of 458 gene expression - DNA methylation
associations as reported in Shakhbazov et al\textsuperscript{40}. These
were filtered to be located on the same chromosome, have robust
correlations after correcting for multiple testing, and to share a SNP
that had a robust cis-acting effect on both the DNA methylation probe
and the gene expression probe. Because only summary statistics were
available (effect, standard error, effect allele, sample size, p-values)
for the instrumental SNP on the methylation and gene expression levels,
the Steiger test of two independent correlations was used to infer the
direction of causality for each of the associations. The Wald ratio test
was then used to estimate the causal effect size for the estimated
direction for each association.

@@

\subsubsection{The causal relationship between gene expression and DNA
methylation
levels}\label{the-causal-relationship-between-gene-expression-and-dna-methylation-levels}

We used the Steiger test to infer the direction of causality between DNA
methylation levels and gene expression levels and found that the causal
direction commonly goes in both directions (Figure 6a), but assuming no
or equal measurement error DNA methylation levels were the predominant
causal factor (\(p = 1.3\times 10^{-5}\)). We then went on to predict
the causal directions of the associations for varying levels of possible
measurement error. Figure 6a shows that the qualitative result of which
of DNA methylation or gene expression is more commonly the causal factor
is very strongly determined by measurement error.

We performed two sample MR\textsuperscript{28} for each association in
the direction of causality that we estimated to be the correct direction
from the Steiger test. We observed that the sign of the MR estimate was
generally in the same direction as the Pearson correlation coefficient
reported by Shakhbazov et al\textsuperscript{40} (Figure 6b). There was
a moderate correlation between the absolute magnitudes of the causal
correlation and the observational Pearson correlation (r = 0.45).
Together these inferences suggest that even for relatively low level
associations it is important to use causal inference methods over
observational associations to infer causal effect sizes.

We also observed that for associations where methylation caused gene
expression the causal effect was more likely to be negative than for the
associations where gene expression caused methylation (OR = 0.61 (95\%
CI 0.36 - 1.03), Figure 6c), suggesting that reducing gene expression
levels at a controlling CpG typically leads to increased gene expression
levels, consistent with expectation\textsuperscript{41}.

@@

Researchers are often confronted with the problem of making causal
inferences using a statistical framework on observational data. In the
epidemiological literature issues of measurement error in mediation
analysis is relatively well explored\textsuperscript{42}. Our analysis
extends this to related methods such as CIT that are widely used in
predominantly 'omic data. These methods are indeed liable to the same
issues as standard mediation based analysis, and specifically we show
that as measurement error in the exposure variable increases, CIT is
likely to have reduced statistical power, and liable to infer the wrong
direction of causality. We also demonstrate that, though unintuitive,
increasing sample size does not resolve the issue, rather it leads to
more extreme p-values for the model that predicts the wrong direction of
causality.

Under many circumstances a practical solution to this problem is to use
Mendelian randomisation instead of methods such as the CIT or similar
that are based on mediation. Inferring the existence of causality using
Mendelian randomisation is robust in the face of measurement error and,
if the researcher has knowledge about the biology of the instrument
being used in the analysis, can offer a direct solution to the issues
that the CIT faces. This assumption is often reasonable, for example
SNPs are commonly used as instruments when they are found in genes with
known biological relevance for the trait of interest. But on many
occasions, especially in the realm of 'omic data, this is not the case,
and methods based on mediation in have been valuable in order to be able
to both ascertain if there is a causal association and to infer the
direction of causality. Here we have described a simple extension to MR
which can be used as an alternative to or in conjunction with mediation
based methods. We show that this method is still liable to measurement
error, but because it has different properties to the CIT it offers
several main advantages. First, it uses a formal statistical framework
to test for the robustness of the assumed direction of causality.
Second, after testing in a comprehensive range of scenarios the MR based
approach is less likely to infer the wrong direction of causality
compared to CIT, while substantially improving power over CIT in the
cases where \(d > 0\).

We demonstrate this new method by evaluating the causal relationships of
458 known associations between DNA methylation and gene expression
levels using summary level data. The causal direction is heavily
influenced by how much measurement error is present in the different
assaying platforms. For example, if DNA methylation measures typically
have higher measuremet error than gene expression measures then our
analysis suggests that DNA methylation levels would be more often the
causal factor in the association. Indeed, previous studies which have
evaluated measurement error in these platforms do support this
position\textsuperscript{43,44}, though making strong conclusions for
this analysis is difficult because measurement error is likely to be
study specific.

In our simulations we focused on the simple case of a single instrument
in a single sample setting with a view to making a fair comparison
between MR and the various mediation-based methods available. However,
MR strategies exist for other scenarios, for example if there are
multiple instruments for the hypothesised exposure then the problems
described here may be circumvented\textsuperscript{16}. Likewise, if
there is at least one instrument for each trait then bi-directional MR
can offer solutions to inferring the causal
direction\textsuperscript{45}. In this work we assumed that pleiotropy
(the influence of the instrument on the outcome through a mechanism
other than the exposure) was not present. Recent method developments in
MR\textsuperscript{24,25} have focused on accounting for the issues that
horizontal pleiotropy can introduce, but how they perform in the
presence of measurement error remains to be explored. An important
advantage that MR confers over most mediation based analysis is that it
can be performed in two samples, which can considerably improve power
and expand the scope of analysis. However, whether there is a
substantive difference in two sample MR versus one sample MR in how
measurement error has an effect is not yet fully understood. We have
also assumed no measurement error in the genetic instrument, which is
not unreasonable given the strict QC protocols that ensure high quality
genotype data is available to most studies. We have restricted the scope
to only exploring non-differential measurement error and avoided the
complications incurred if measurement error in the exposure and outcome
is correlated, however our analysis goes beyond many previous
explorations of measurement error by assessing the impacts of both
imprecision (noise) and linear transformations of the true variable on
causal inference.

Mediation based network approaches, that go beyond analyses of two
variables, are very well established\textsuperscript{33} and have a
number of extensions that make them valuable tools, including for
example network construction. But because they are predicated on the
basic underlying principles of mediation they are liable to suffer from
the same issues of measurement error. Recent advances in MR methodology,
for example applying MR to genetical genomics\textsuperscript{46},
multivariate MR\textsuperscript{47} and mediation through
MR\textsuperscript{48--50} may offer more robust alternatives for these
more complicated problems.

@@

Figure 1: Gene expression levels (blue blocks) and DNA methylation
levels (green triangles) may be correlated but the causal structure is
unknown. If a SNP (yellow circle) is associated with both DNA
methylation and gene expression levels then it can be used as an
instrument, but there are three basic competing models for these
variables. The causal inference test (CIT) attempts to delineate between
them. a) Methylation causes gene expression. The left figure shows that
the SNP influences methylation levels that in turn influence gene
expression levels. The right figure shows the directed acyclic graph
that represents this model. Faded symbols represent the measured values
whereas solid symbols represent the true values. b) The same as in A,
except the causal direction is from gene expression to DNA methylation.
c) A model of confounding, where gene expression and DNA methylation are
not causally related, but the SNP influences them each through separate
pathways or a confounder.

@@

Figure 2: The CIT was performed on simulated variables where the
exposure influenced the outcome and the exposure was instrumented by a
SNP. The test statistic from CIT when testing if the exposure caused the
outcome (the true model) is in red, and the test for the outcome causing
the exposure (false model) is in green. Rows of plots represent the
sample sizes used for the simulations. As measurement imprecision
increases (decreasing values on x-axis) the test statistic for the
incorrect model gets stronger and the test statistic for the correct
model gets weaker.

@@

Figure 6: Using 458 putative associations between DNA methylation and
gene expression we used the Steiger test to infer the direction of
causality between them. a) The rightmost bar shows the proportion of
associations for each of the two possible causal directions (colour key)
assuming no measurement error in either gene expression or DNA
methylation levels. The proportions change when we assume different
levels of measurement error in gene expression levels (x-axis) or DNA
methylation levels (columns of boxes). If there is systematically higher
measurement error in one platform than the other it will appear to be
less likely to be the causal factor. b) The relationship between the
Pearson correlation between DNA methylation and gene expression levels
(x-axis) and the causal estimate (scaled to be in standard deviation
units, y-axis). c) Distribution of estimated causal effect sizes,
stratified into associations inferred to be due to DNA methylation
causing expression (blue) and expression causing DNA methylation (red).

@@

\hypertarget{ref-Pierce2013}{}
28. Pierce, B. L. \& Burgess, S. Efficient design for Mendelian
randomization studies: subsample and 2-sample instrumental variable
estimators. \emph{American journal of epidemiology} \textbf{178,}
1177--84 (2013).

\hypertarget{ref-Ashenfelter1994}{}
29. Ashenfelter, O. \& Krueger, A. B. Estimates of the Economic Return
to Schooling from a New Sample of Twins. \emph{The American Economic
Review} \textbf{84,} 1157--1173 (1994).

\hypertarget{ref-Nagarajan2013}{}
30. Nagarajan, R. \& Scutari, M. Impact of noise on molecular network
inference. \emph{PloS one} \textbf{8,} e80735 (2013).

\hypertarget{ref-Shpitser2010}{}
31. Shpitser, I., VanderWeele, T. \& Robins, J. On the validity of
covariate adjustment for estimating causal effects. \emph{Proceedings of
the Twenty Sixth Conference on Uncertainty in Artificial Intelligence
(UAI-10)} 527--536 (2010).

\hypertarget{ref-Wang2015}{}
32. Wang, L. \& Michoel, T. Detection of regulator genes and eQTLs in
gene networks. \emph{arXiv} \textbf{arXiv:1512,} (2015).

\hypertarget{ref-Lagani2015}{}
33. Lagani, V., Triantafillou, S., Ball, G., Tegner, J. \& Tsamardinos,
I. in \emph{Uncertainty in biology: A computational modeling approach}
47 (Springer, 2015). at
\textless{}\href{https://books.google.com/books?id=8SLUCgAAQBAJ\%7B/\&\%7Dpgis=1}{https://books.google.com/books?id=8SLUCgAAQBAJ\{\textbackslash{}\&\}pgis=1}\textgreater{}

\hypertarget{ref-Lawlor2017}{}
34. Lawlor, D. A., Tilling, K. \& Davey Smith, G. Triangulation in
aetiological epidemiology. \emph{International Journal of Epidemiology}
\textbf{19,} dyw314 (2017).

@@

\hypertarget{ref-Shakhbazov2016}{}
40. Shakhbazov, K. \emph{et al.} Shared genetic control of expression
and methylation in peripheral blood. \emph{BMC genomics} \textbf{17,}
278 (2016).
